# Reproducible Research: Peer Assessment 1

This work is part of the 1st Peer Assessment for the Reproducible Research Course. The first step in this work is to obtain the data set whic his available here [link](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)

## Loading and preprocessing the data

For this part of the assignment the code below does the following:
- Unzip the zipped file
- Load data in R and define the appropriate classes for each field in the data set
- Obtain  the column names (colNames), number of rows in the data set (numberofRows)
- Print numberofRows and colNames
- Print the time/date the dataset was downloaded from the github

```{r,echo=TRUE}

sourceFile <- "activity.zip"

## show the time the file was downloaded
dateaccessed <-as.character(file.info(sourceFile)$mtime)
cat("activity.zip was downloaded on: ", dateaccessed)

## load the file into R and extract the activity.csv from the zip file      
dataSet <- read.csv(unz(sourceFile, "activity.csv"), header=TRUE, colClasses=c("numeric", "Date", "numeric"), stringsAsFactors=FALSE)
## get the column names in the data set
colNames <- colnames(dataSet)
# number of rows in the dataset
numberofRows <- nrow(dataSet)
## print colNames and numberofRows
cat("Number of Records in the Data Set: ", numberofRows, "\n","Columns Names in the Data Set:", colNames)

```


## What is mean total number of steps taken per day?

To get the mean, first we construct a subset of data that contains the number of steps per date:
```{r, echo=TRUE}
## {plyr} is required before calling the ddply function 
require(plyr)
## for this part, remove the missing values 
dataSubset <- ddply(dataSet,.(date), summarize, steps=sum(steps, na.rm=TRUE))
hist(dataSubset$steps, breaks=15, main ="Number of Steps Taken per Day",
             col="seagreen", xlab ="Steps per Day", ylab= "Frequency")
```
To calculate the mean and median, the dataSubset will be used. 
```{r,echo=TRUE}
meanVal <- mean(dataSubset$steps)
medVal <- median(dataSubset$steps)
##show the results 
cat("Mean value of steps taken per day is: ", meanVal, "\nMedian value of steps taken per day is: ", medVal)
```
## What is the average daily activity pattern?

- To make the time seriese, we need to take the average number of steps accross all available intervals
- This requires creating a subset grouped by interval as follows: 
```{r,echo=TRUE}
## create a subset i.e. dataframe of average number of steps grouped by interval
## overall there is 288 interval (i.e. nrow(dataSubset) results in 288)
dataSubset <- ddply(dataSet,.(interval), summarize, steps=mean(steps, na.rm=TRUE))

## plot a time series of average number of taken steps per activity
plot(dataSubset$interval, dataSubset$steps, 
             type="l", xlab="Five minutes intervals", ylab="Average take steps")

```

The code below reports which 5-minute interval, on average across all the days in the dataset,
contains the maximum number of steps:

```{r,echo=TRUE}
maxInterval <- dataSubset[which.max(dataSubset[,"steps"]),]
## print the results
cat("The Interval ", maxInterval$interval, " On average accross all days, has the maximum number of steps which is equal to", maxInterval$steps)
```
## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
